---
title: "Customer Churn Prediction Group Project - Random Forest"
author: "Thanh Tran"
date: "2025-04-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##1. Loading packages: loading caret (predictive modeling), dplyr (data wrangling), ranger (random forest algorithm)
```{r}
library(caret)
library(dplyr)
library(ranger)
```

##2. Importing data
```{r}
ecommercechurn <- read.csv("C:/Users/ThanhT24/Downloads/OIM 454/E Commerce Dataset For Final Project.csv")
```

##3. Exploring data
```{r}
head(ecommercechurn)
nrow(ecommercechurn)
```

##4. Pre-process data for binary classification
```{r}
ecommercechurn$Churn <- factor(ecommercechurn$Churn, levels = c(0, 1))
head(ecommercechurn)
```

##5. Partition training and validation sets: 80% of data randomly allocated to training
80% of the data (3,220 records) goes into the training dataset, while the remaining 20% (805 records) to validation and test from the original dataset consists of 4,025 records.
```{r}
sample <- sample.int(n = nrow(ecommercechurn), size = nrow(ecommercechurn)*0.80, replace = F)

ecommercechurn_training <- ecommercechurn[sample, ] ##Yields training dataset that is the training percentage % 

ecommercechurn_validationtest <- ecommercechurn[-sample, ] ##Yields validation dataset
```
```{r}
nrow(ecommercechurn_training)
nrow(ecommercechurn_validationtest)
```
Among the 20%, 10% to the validation set, which is equivalent to 402 records, and the remaining 10% consists of 403 records to the test set.
```{r}
sample <- sample.int(n = nrow(ecommercechurn_validationtest), size = nrow(ecommercechurn_validationtest)*0.50, replace = F) ##Validation percentage = what percentage of this validation + test block should go into validation
ecommercechurn_validation <- ecommercechurn_validationtest[sample, ] ##Yields validation dataset
ecommercechurn_test <- ecommercechurn_validationtest[-sample, ] ##Yields test portion
```
```{r}
nrow(ecommercechurn_validation)
nrow(ecommercechurn_test)
```

##6. Train random forest model
Using non-normalized data to training RF model. Only necessary to normalize data for kNN in this context.
```{r}
rf_fit <- train(Churn ~ ., data = ecommercechurn_training, method = "ranger") ##Does not need to use normalized data so using original training partition
rf_fit
```

##7. Generate random forest predictions on validation data and test data (data does not need to be normalized)
Making predictions on non-normalized validation data.
```{r}
rf_validation_predictions <- predict(rf_fit, ecommercechurn_validation)
```
```{r}
rf_test_predictions <- predict(rf_fit, ecommercechurn_test)
```
The model performed strongly on both validation and test sets. On the validation set, it achieved 98.51% accuracy, 93.55% sensitivity, 99.41% specificity, and 96.67% precision. On the test set, accuracy, specificity, and precision decreased slightly to 97.52%, 98.15%, and 92.50%, respectively, while sensitivity improved to 94.87%. Overall, the results indicate that the model generalizes well and remains highly effective at identifying both churned and non-churned customers.
```{r}
confusionMatrix(rf_validation_predictions, ecommercechurn_validation$Churn,positive="1") ##Random forest validation predictions 
```

```{r}
confusionMatrix(rf_test_predictions, ecommercechurn_test$Churn,positive="1") ##Random forest test predictions 
```